\documentclass{article}

\usepackage{amssymb}
\usepackage[T1]{fontenc}
% \usepackage[utf8]{inputenc}
\usepackage[greek,polish,english]{babel}

% Set page size and margins
\usepackage[a4paper,top=1cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{csquotes}
% \usepackage{txfonts}
% \usepackage{pxfonts}
% \usepackage{newtxmath}
\usepackage{mathtools}
\usepackage{wasysym}
\usepackage{xspace}

\usepackage{amsthm}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{cleveref}
\usepackage{textcomp}
\usepackage[multiple, perpage]{footmisc}
\usepackage{lipsum}
\usepackage{sidenotes}
\usepackage{tikz,tikz-3dplot}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{framed}
\usepackage{amsmath}
\usepackage{xcolor, colortbl}
\usepackage{csquotes}
\usepackage[shortlabels]{enumitem}
\definecolor{c1}{HTML}{377eb8}
\definecolor{c2}{HTML}{ff7f00}
\definecolor{c3}{HTML}{4daf4a}
\definecolor{c4}{HTML}{d62728}
\usetikzlibrary{arrows}
\usetikzlibrary{patterns}
\usetikzlibrary{hobby}

\usetikzlibrary{backgrounds}
\tdplotsetmaincoords{80}{45}
\tdplotsetrotatedcoords{-90}{180}{-90}
\tikzset{surface/.style={draw=red!70!black, fill=red!40!white, fill opacity=.6}}

\renewcommand*{\thefootnote}{\alph{footnote}}
\interfootnotelinepenalty=1000000000

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\AfterEndEnvironment{theorem}{\noindent\ignorespaces}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\AfterEndEnvironment{definition}{\noindent\ignorespaces}

\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\AfterEndEnvironment{example}{\noindent\ignorespaces}

\newcommand*{\Ts}{T^*}
\newcommand*{\id}{\equiv}
\newcommand*{\ra}{\rightarrow}

\newcommand*{\V}{\texttt{V}}
\newcommand*{\FOR}{\texttt{FOR}}
\newcommand*{\FORx}{\texttt{FOR}^X}
\newcommand*{\ID}{\texttt{ID}}
\newcommand*{\IDx}{\texttt{ID}^X}
\newcommand*{\SUB}{\texttt{SUB}}
\newcommand{\SCI}{$\mathsf{SCI}$\xspace}
\newcommand{\PC}{$\mathsf{PC}$\xspace}
\newcommand{\M}{\mathcal{M}\xspace}

\makeatletter
\newcommand{\leqnomode}{\tagsleft@true\let\veqno\@@leqno}
\newcommand{\reqnomode}{\tagsleft@false\let\veqno\@@eqno}
\makeatother

\title{Decision procedures for a non-Fregean logic \SCI}
\author{Adrian Siwiec}

\begin{document}
\date{\today}

\maketitle

Test: ążźćłóęń

\section{Sentential logic and the Fregean axiom}
\section{\SCI}
\subsection{Basic notions}

\begin{definition}[Vocabulary of \SCI]
    The vocabulary of \SCI consists of symbols from the following pairwise disjoint sets:
    \begin{itemize}
        \item $\V = \{p, q, r, ...\}$ -- a countable infinite set of propositional variables,
        \item $\{\lnot, \ra, \id \}$  -- the set consisting of the unary operator of negation ($\lnot$) and binary operators of implication ($\ra$) and identity ($\id$),
        \item $\{(,)\}$ -- the set of auxiliary symbols.
    \end{itemize}
\end{definition}
\begin{definition}[Formulas of \SCI]
    The set of formulas of \SCI is defined with the following grammar:
    $$
        \FOR \ni \varphi ::= p  \; | \;  \lnot \varphi \; | \; (\varphi \ra \varphi) \; | \; (\varphi \id \varphi)
    $$ where $p \in \V$ is a propositional variable.
\end{definition}
The propositional variables will also be called atomic formulas.

From now on whenever we write $p, q, r, ... $ we will mean the atomic formulas.
We will omit brackets when it will lead to no misunderstanding. We will write
$\varphi \not \id \psi$ as a shorthand for $\lnot(\varphi \id \psi)$.

The set of identities $\ID$ is a set of formulas $\varphi \id \psi$ where
$\varphi, \psi \in \FOR$. Formulas $\varphi \id \varphi$ are the trivial
identities.

\begin{definition}[Subformulas]
    For a formula $\varphi \in \FOR$ let us define the set of subformulas of $\varphi$ as:
    $$
        \SUB(\varphi) = \begin{cases}
            \{p\},                                            & \text{if } \varphi = p \in \V,                                                   \\
            \{\varphi\} \cup \SUB(\psi),                      & \text{if } \varphi = \lnot \psi,                                                 \\
            \{\varphi\} \cup \SUB(\psi) \cup \SUB(\vartheta), & \text{if } \varphi = \psi \ra \vartheta, \text{or } \varphi = \psi \id \vartheta \\
        \end{cases}
    $$
\end{definition}
By $\varphi(\psi / \vartheta)$ we will denote the formula $\varphi$ with all
occurrences of its subformula $\psi$ substituted with $\vartheta$.

\begin{definition}[Simple formulas]
    The formula $\varphi$ is called a simple formula if it has one of the following form:
    $$
        p,\ \lnot p,\ p \id q,\ p \not \id q,\ p \id \lnot q,\ p \id (q \ra r),\ p \id (q \id r)
    $$
    for $p, q, r \in \V$.
\end{definition}

\begin{definition}[Size of a formula]
    Given a formula $\varphi$, let us define its size $s(\varphi)$:
    $$
        s(\varphi) = \begin{cases}
            1,                          & \text{if } \varphi = p,                                                                      \\
            s(\psi) + 1,                & \text{if } \varphi = \lnot \psi,                                                             \\
            s(\psi) + s(\vartheta) + 1, & \text{if } \varphi = \psi \equiv \vartheta,\text{ or } \varphi = \psi \rightarrow \vartheta.
        \end{cases}
    $$
\end{definition}
$\V$ is a countable set. Let us take any full ordering of it and mark it as $<$. Let us then
extend it by saying that for each $p \in \V$: $p < \lnot <\xspace (\ < \ )\xspace<\ \ra\ <\ \id$. Now,
let us define a ordering of formulas $\prec$ to be a lexicographical ordering
with $<$.

If we consider formulas that contain only the negation and implication
operators, they form a classical Propositional Calculus. For simplicity, in
\SCI we'll consider every tautology of the classical Propositional Calculus to
be an axiom.

\begin{definition}[Axiomatization of \SCI]
    \label{axiomatization}
    \SCI is axiomatized with the following axioms:
    \begin{itemize}
        \item Any tautology of the classical Propositional Calculus
        \item ($Ax1$) $\varphi \id \varphi$,
        \item ($Ax2$) $(\varphi \id \psi) \ra (\varphi \ra \psi)$,
        \item ($Ax3$) $(\varphi \id \psi) \ra (\lnot \varphi \id \lnot \psi)$,
        \item ($Ax4$) $(\varphi \id \psi) \ra ((\chi \id \theta) \ra ((\varphi \ra \chi) \id (\psi \ra \theta)))$,
        \item ($Ax5$) $(\varphi \id \psi) \ra ((\chi \id \theta) \ra ((\varphi \id \chi) \id (\psi \id \theta)))$.
    \end{itemize}
    The only inference rule is the \emph{modus ponens} rule:
    $$
        \mathsf{MP}: \frac{
            \varphi, \ \varphi \ra \psi}%
        { \psi }
    $$
\end{definition}

\begin{definition}[A thesis of \SCI]
    A formula $\varphi$ is \emph{a thesis of \SCI} if there exists a finite sequence of formulas $\varphi_1, ..., \varphi_n$ ($n \geq 1$), such that $\varphi = \varphi_n$, and for all $i \in \{1, ..., n\}$ the formula $\varphi_i$ is either an axiom of \SCI, or it is inferred from formulas $\varphi_j$, $\varphi_k$ ($j, k <i$) via the \emph{modus ponens} rule. If $\varphi$ is a thesis of \SCI we will denote it by $\vdash \varphi$, say that $\varphi$ is \emph{provable in \SCI} and call the sequence $\varphi_1, ..., \varphi_n$ the \emph{proof of $\varphi$}.
    \label{thesis}
\end{definition}

Let us now give some semantic definitions:

\begin{definition}[\SCI-model]
    A model of \SCI (or an \SCI-model) model is a structure $\mathcal{M} = (U, D, \tilde{\lnot}, \tilde{\ra}, \tilde{\id})$ where:
    \begin{itemize}
        \item $U \not = \emptyset$ is an \emph{universe} of $M$,
        \item $\emptyset \not = D \subsetneq U$ is a \emph{set of designated values},
        \item $\tilde{\lnot}$ is an unary operation on $U$, such that for all $a \in U$: $\tilde{\lnot}a \in D$ if and only if $a \not \in D$,
        \item $\tilde{\ra}$ is a binary operation on $U$, such that for all $a, b \in U$: $a \tilde{\ra} b \in D$ if and only if $a \not \in D$ or $b \in D$,
        \item $\tilde{\id}$ is a binary operation on $U$, such that for all $a, b \in U$: $a \tilde{\id} b \in D$ if and only if $a = b$ (that is, if and only if $a$ and $b$ denote the same element of the universe $U$).
    \end{itemize}
    \label{model}
\end{definition}

If an universe $U$ is finite, we'll call a given \SCI-model a \emph{finite
    \SCI-model}.

\begin{definition}[Valuation]
    Given an \SCI-model $\mathcal{M} = (U, D, \tilde{\lnot}, \tilde{\ra}, \tilde{\id})$, a valuation in $\mathcal{M}$ is a function $V : \FOR \longrightarrow U$ that assigns a value $V(p) \in U$ for all propositional variables $p$, and such that for all $\varphi, \psi \in \FOR$:
    \begin{itemize}
        \item $V(\lnot \varphi) = \tilde{\lnot}V(\varphi)$
        \item $V(\varphi \ra \psi) = V(\varphi) \tilde{\ra} V(\psi)$
        \item $V(\varphi \id \psi) = V(\varphi) \tilde{\id} V(\psi)$
    \end{itemize}
    If $V(\varphi) = a$ we will call $a$ the \emph{denotation of $\varphi$}.
    \label{valuation}
\end{definition}

\begin{definition}[Satisfaction of a formula]
    Given an \SCI-model $\mathcal{M} = (U, D, \tilde{\lnot}, \tilde{\ra}, \tilde{\id})$ and a valuation $V$ in $\mathcal{M}$, a formula $\varphi$ is \emph{satisfied in $\mathcal{M}$ by $V$} if and only if $V(\varphi) \in D$. If a formula $\varphi$ is satisfied in $\mathcal{M}$ by a valuation $V$, then will denote it by $M, V \models \varphi$.
\end{definition}

\begin{definition}[Truth of a formula]
    Given an \SCI-model $\mathcal{M} = (U, D, \tilde{\lnot}, \tilde{\ra}, \tilde{\id})$, a formula is \emph{true in $\mathcal{M}$} if and only if it is satisfied in $\mathcal{M}$ by all valuations of $\mathcal{M}$. If a formula $\varphi$ is true in $\mathcal{M}$, then we will denote it by $\mathcal{M} \models \varphi$.
\end{definition}

\begin{definition}[Validity of a formula]
    A formula is \emph{valid in \SCI} (or \emph{\SCI-valid}) if and only if it is true in all models. If a formula $\varphi$ is \SCI-valid we will denote it by $ \models \varphi$.
\end{definition}

Let us see some examples:

TODO

\begin{theorem}[Soundness of \SCI]
    For every \SCI-formula $\varphi$, if $\varphi$ is provable in \SCI, then $\varphi$ is valid in \SCI.
    \label{soundness}
\end{theorem}
\begin{proof}
    The proof will have two main parts:
    \begin{enumerate}[(a)]
        \item Every axiom of \SCI is valid in \SCI.
        \item By applying the \emph{modus ponens} rule to valid formulas $\varphi$ and
              $\varphi \ra \psi$ the inferred formula $\psi$ is also valid.
    \end{enumerate}

    Let us show (a) first:
    \begin{itemize}
        \item First, we need to prove that every tautology of the classical Propositional
              Calculus (\PC) is valid in \SCI.

              Let $\varphi$ be a tautology of \PC. Let us take any \SCI-model $\mathcal{M}$
              and any \SCI-valuation $\mathcal{M}, V$ in it. We want to show that
              $\mathcal{M}, V \models \varphi$.

              Given $\mathcal{M}, V$, we will construct $V'$ that will be a \PC valuation of
              $\varphi$. We can do this, because $\varphi$ doesn't contain operator $\id$.

              For every $\psi \in \SUB(\varphi)$, let us define:

              $$
                  V'(\psi) = \begin{cases}
                      1,                           & \text{if } \psi = p \in \V \text{ and } V(p) \in D,      \\
                      0,                           & \text{if } \psi = p \in \V \text{ and } V(p) \not \in D, \\
                      1-V'(\chi),                  & \text{if } \psi = \lnot \chi,                            \\
                      max(1-V'(\chi), V'(\theta)), & \text{if } \psi = \chi \ra \theta.                       \\
                  \end{cases}
              $$

              It is easy to see that for every $\psi \in \SUB(\varphi)$, $V'(\psi) = 1$ if
              and only if $V(\psi) \in D$.

              The $V'$ function is constructed in the same way the valuation function in the
              \PC is constructed, therefore since $\varphi$ is a tautology of \PC, we have
              that $V'(\varphi) = 1$. So, we have that $V(\varphi) \in D$, whis is what we
              wanted to show.

        \item Second, we want to show that axioms $(Ax1) - (Ax5)$ are valid in \SCI.

              \begin{itemize}
                  \item $(Ax1)\  \varphi \id \varphi$

                        Let us take any \SCI model and valuation $\mathcal{M}, V$. Based on
                        \cref{valuation}, we have that $V(\varphi \id \varphi) = V(\varphi) \tilde{\id}
                            V(\varphi)$. Based on definition of $\tilde{\id}$ from \ref{model}, we have
                        that $V(\varphi) \tilde{\id} V(\varphi)$ is in $D$ if and only if $V(\varphi) =
                            V(\varphi)$, which is trivially the case. So, we have that $\mathcal{M}, V
                            \models \varphi \id \varphi$.
                  \item $(Ax2)\ (\varphi \id \psi) \ra (\varphi \ra \psi)$

                        Let us take any \SCI model and valuation $\mathcal{M}, V$. Based on
                        \cref{valuation}, we have that $V((\varphi \id \psi) \ra (\varphi \ra \psi)) =
                            V(\varphi \id \psi) \tilde{\ra} V(\varphi \ra \psi)$. There are two cases.

                        \begin{enumerate}
                            \item[1°] $V(\varphi \id \psi) \not \in D$. Then, by definition of $\tilde{\ra}$ in \ref{model}, we have that $V(\varphi \id \psi) \tilde{\ra} V(\varphi \ra \psi) \in D$, so $\mathcal{M}, V \models (\varphi \id \psi) \ra (\varphi \ra \psi)$.
                            \item[2°] $V(\varphi \id \psi) \in D$. By \cref{valuation} it means, that $V(\varphi) \tilde{\id} V(\psi)$, which by \cref{model} means that $V(\varphi) = V(\psi)$.

                                By \cref{valuation} $V(\varphi \ra \psi)$ is equal to $V(\varphi) \tilde{\ra}
                                    V(\psi)$. By \cref{model} $V(\varphi) \tilde{\ra} V(\psi)$ is in $D$ if and
                                only if $V(\varphi) \not \in D$ or $V(\psi) \in D$, but since $V(\varphi) =
                                    V(\psi)$ we have that $V(\varphi) \tilde{\ra} V(\psi)$ is in $D$ if and only if
                                $V(\varphi) \not \in D$ or $V(\varphi) \in D$, which is trivially the case.
                        \end{enumerate}

                        So, we have that $\mathcal{M}, V \models (\varphi \id \psi) \ra (\varphi \ra
                            \psi)$.

                  \item Validity of axioms $(Ax3), (Ax4)$ and $(Ax5)$ can be shown in a similar way.
                        (TODO: a może rozpisać dla jasności?)
              \end{itemize}
    \end{itemize}

    To show (b), let take any valid formulas $\varphi$ and $\varphi \ra \psi$ and
    any \SCI model and valuation $\mathcal{M}, V$. From validity of $\varphi \ra
        \psi$, we know that $V(\varphi \ra \psi) \in D$. From \cref{valuation} we have
    that $D \ni V(\varphi \ra \psi) = V(\varphi) \tilde{\ra} V(\psi)$. From
    \cref{model}, since we know that $V(\varphi) \tilde{\ra} V(\psi)$ we know that
    $V(\varphi) \not \in D$, or $V(\psi) \in D$. But we have that $\varphi$ is
    valid, so $V(\varphi) \in D$, so it must be that $V(\psi) \in D$, so
    $\mathcal{M}, V \models \psi$. The same holds for any other \SCI model and
    valuation, so $\models \psi$, which is what we wanted to show.

    Now, looking at \cref{thesis}, for a given provable $\varphi$, let us take its
    proof $\varphi_1, ..., \varphi_n = \varphi$. Every subsequent formula in this
    proof is either an \SCI axiom and thus, by (a), valid, or is inferred by the
    \emph{modus ponens} rule from valid formulas and thus, by (b) valid. So,
    $\varphi$ is valid.
\end{proof}

\begin{theorem}[Completeness of \SCI]
    For every \SCI-formula $\varphi$, if $\varphi$ is valid in \SCI, then $\varphi$ is provable in \SCI.
\end{theorem}
\begin{proof}
    To show that every valid \SCI-formula is \SCI-provable is, by contraposition, to show that every non-\SCI-provable formula is non-\SCI-valid.

    To do this, we will construct a \SCI-model and a valuation in which every
    \SCI-provable formula is satisfied (this model will be similar in style to the
    Herbrandt model). Then, it will be shown that for every formula that is not
    \SCI-provable, this model and this valuation will be a witness of its
    non-\SCI-validity, thus proving completeness of \SCI

    {\color{red}Wydaje mi się, że pomieszałem coś w notatkach z tego dowodu. Na wykładzie co prawda dla każdego niesprzecznego zbioru formuł $X$ konstruowaliśmy model, w którym były prawdziwe wszystkie formuły $X \vdash \varphi$, ale dla samego dowodu pełności i tak stawialiśmy na koniec (chyba?) $X = \emptyset$, wiec uprościłem.

        Próbowałem to porównać z [Suszko, Bloom 1972], ale tam ten dowód jest bardzo
        lakonicznie opisany.

        A może bierze się niedowodliwą formułę i dla niej konkretnie buduje model (lub
        wartościowanie), które ją inwaliduje?}

    First, let us define a relation $R$ on \SCI-formulas: $$ \varphi R \psi \text{
            if and only if} \vdash \varphi \id \psi $$

    Note that for clarity we're using infix notation for relations. We want to show
    that $R$ is an equivalence relation.
    \begin{enumerate}[(a)]
        \item $R$ is reflexive. It is trivially true from axiom $(Ax1)$.
        \item $R$ is symmetrical, that is for all $\varphi, \psi \in \FOR, \varphi R \psi$ if and only if $\psi R \varphi$.

              As we have shown in example TODO, we can prove that for all $\varphi, \psi$,
              $\vdash (\varphi \id \psi) \ra (\psi \id \varphi)$. If $\varphi R \psi$, then
              $\vdash \varphi \id \psi$, so by \emph{modus ponens} we can infer $\vdash \psi
                  \id \varphi$, so $\psi R \varphi$. In the same way we can show that if $\psi R
                  \varphi$ then $\varphi R \psi$, so $R$ is symmetrical.
        \item $R$ is transitive, that is for all $\varphi, \psi, \vartheta \in \FOR$, if $\varphi R \psi$ and $\psi R \vartheta$ then $\varphi R \vartheta$.

              As we have shown in example TODO, we can prove that for all $\varphi, \psi,
                  \vartheta \in \FOR$, $\vdash (\varphi \id \psi) \ra ((\psi \id \vartheta) \ra
                  (\varphi \id \vartheta))$. Let us assume that $\varphi R \psi$ and $\psi R
                  \vartheta$. This gives us that $\vdash \varphi \id \psi$ and $\vdash \psi \id
                  \vartheta$. Applying \emph{modus ponens} twice we get $\vdash \varphi \id
                  \vartheta$, so $\varphi R \vartheta$.
    \end{enumerate}

    So, $R$ is an equivalence relation. Let us now show that it is a congruence,
    that is:
    \begin{enumerate}[(a)]
        \item For all $\varphi, \psi \in \FOR$, if $\varphi R \psi$, then $\lnot \varphi R
                  \lnot \psi$. This is trivially true from axiom $(Ax3)$.
        \item For all $\varphi, \psi, \vartheta, \chi \in \FOR$, if $\varphi R \psi$ and
              $\vartheta R \chi$, then $\varphi \ra \vartheta R \psi \ra \chi$. This is
              easily proven from $(Ax4)$.
        \item For all $\varphi, \psi, \vartheta, \chi \in \FOR$, if $\varphi R \psi$ and
              $\vartheta R \chi$, then $\varphi \id \vartheta R \psi \id \chi$. This is
              easily proven from $(Ax5)$.
    \end{enumerate}

    Now that we know that $R$ is a congruent equivalence relation, we can define a
    structure: $\mathcal{M} = (U, D, \tilde{\lnot}, \tilde{\ra}, \tilde{\id})$,
    such that:
    \begin{itemize}
        \item $U = \{ |\varphi|_R : \varphi \in \FOR \}$ (by $|\varphi|_R$ we mean an equivalence class of $R$ of which $\varphi$ is a member),
        \item $D = \{ |\varphi|_R : \varphi \in \FOR, \vdash \varphi\}$
        \item $\tilde{\lnot}$ is an unary function on equivalence classes of $R$, such that $\tilde{\lnot}|\varphi|_R = |\lnot \varphi|_R$
        \item $\tilde{\ra}$ is a binary function on equivalence classes of $R$, such that $|\varphi|_R \tilde{\ra} |\psi|_R = |\varphi \ra \psi|_R$
        \item $\tilde{\id}$ is a binary function on equivalence classes of $R$, such that $|\varphi|_R \tilde{\id} |\psi|_R = |\varphi \id \psi|_R$
    \end{itemize}

    Let us add some comments:
    \begin{itemize}
        \item For any \SCI-formulas $\varphi, \psi$, such that $\varphi \id \psi$, $\vdash
                  \varphi$ if and only if $\vdash \psi$. This follows from $(Ax2)$. Because of
              that, $|\varphi|_R = |\psi|_R$ and so $D$ is well defined.
        \item From congruence of $R$, we know that for any $\varphi, \psi \in \FOR$ if
              $|\varphi|_R = |\psi|_R$, then $|\lnot \varphi|_R = |\lnot \psi|_R$, so
              $\tilde{\lnot}$ is a well defined function.
        \item Similarly, form the congruence of $R$ we can see that $\tilde{\ra}$ and
              $\tilde{\id}$ are well defined functions.
    \end{itemize}

    Now we will want to show that $\mathcal{M}$ is an \SCI-model. From the
    \cref{model} we want to show that:
    \begin{itemize}
        \item $U \not = \emptyset$ -- this is trivial,
        \item $\emptyset \not = D$ -- this is true, because for example $\vdash p \id p$
        \item $D \subsetneq U$ --  $D \subseteq U$ is trivial from the definition of $D$. $D \not = U$, because there are formulas which are not valid (e.g. $\lnot(p \id p)$), which from \cref{soundness} are not provable.
        \item For all $a \in U: \tilde{\lnot} a \in D$ if and only if $a \not \in D$

              {\color{red} Czy to nie jest fałsz? Weźmy $a=|p\id q|$.

                      Chcemy pokazać: $\tilde{\lnot}|p \id q| \in D$ iff $|p \id q| \not \in D$.

                      Prawa strona: Prawdziwa, bo $\not \vdash |p \id q|$

                      Lewa strona: $\tilde{\lnot}|p \id q| \xLeftrightarrow{\text{def.}
                              \tilde{\lnot}}$ $|\lnot(p \id q)|$. Ale: $\not \vdash \lnot(p \id q)$, więc
                      $|\lnot(p \id q)| \not \in D$, więc lewa nieprawdziwa.}
        \item For all $a, b \in U: a \tilde{\ra}b \in D$ if and only if $a \not \in D$ or $b
                  \in D$.

              TODO.
        \item For all $a, b \in U: a \tilde{\id}b \in D$ if and only if $a = b$.

              TODO.
    \end{itemize}

    So, $\mathcal{M}$ is a \SCI-model. Now, let us define a valuation $V$ for all
    $\varphi \in \FOR$: $$ V(\varphi) = |\varphi|_R $$

    We need to show that $V$ is indeed a valuation:
    \begin{itemize}
        \item TODO (chyba proste wszystko)
    \end{itemize}

    Now, for the final proof of completeness let us take a non-provable formula
    $\varphi$. Since $\not \vdash \varphi$ we have that $|\varphi|_R \not \in D$,
    so $V(\varphi) \not \in D$. So, we have found a model and a valuation
    $\mathcal{M},V$, such that $\mathcal{M},V \not \models \varphi$, so $\varphi$
    is not valid.

        {\color{red} W notatkach mam:
            \enquote{
                \begin{itemize}
                    \item Mamy dane $X, \varphi$: $X \not \vdash \varphi$
                    \item budujemy model $\M^X, V^X$, t.że $X \vdash \psi \Rightarrow |\psi|_{R_X} \in
                              D^X$
                    \item Ale oczywiście skoro $X \not \vdash \varphi$ to $|\varphi|_{R_X} \not \in D^X$,
                          więc $X \not \models \varphi$
                \end{itemize}}\\
            Więc $X$ nic tutaj nie pomaga, stawiając $X = \emptyset$ dostaję to co jest u góry.
        }

\end{proof}

\begin{definition}[Decidability]
    A logic is \emph{decidable} if there exists an effective method, that is, an algorithm that will always terminate, to determine whether any formula of this logic is a theorem.
\end{definition}
%
We will want to show that \SCI is decidable. To do this, first let us show the
following:
\begin{theorem}[Finite model property of \SCI]
    \label{finite_model}
    For any $\varphi \in \FOR$, if there exists $\mathcal{M},V$, such that $\mathcal{M},V \models \varphi$, then there exists a finite \SCI-model $\mathcal{M'},V'$, such that $\mathcal{M'},V'\models \varphi$.
\end{theorem}
\begin{proof}
    We are given $\mathcal{M}, V$ and $\varphi \in FOR$, such that $\mathcal{M}, v \models \varphi$. Let us define $\mathcal{M'} = (U', D', \tilde{\lnot}', \tilde{\ra}')$ and $V'$:
    \begin{itemize}
        \item $U' = \{v(\psi) : \psi \in \SUB(\varphi)\} \cup \{1, 0\}$ -- let us create $1, 0$, such that $1, 0 \not \in U$,

        \item $D' = \{V(\psi) : \psi \in \SUB(\varphi), V(\psi) \in D\} \cup \{1\}$
        \item $
                  \tilde{\lnot}'a = \begin{cases}
                      \tilde{\lnot}a , & \text{if exists } \psi \text{ such that } a = V(\psi) \text{ and } \lnot \psi \in \SUB(\varphi) \\
                      0,               & \text{otherwise, if } a \in D'                                                                  \\
                      1,               & \text{otherwise, if } a \not \in D'                                                             \\
                  \end{cases}
              $
        \item $
                  a\tilde{\ra}'b = \begin{cases}
                      a\tilde{\ra}b , & \text{if exist } \psi, \vartheta \text{ such that } a = V(\psi), b = V(\vartheta) \text{ and } \psi \ra \vartheta \in \SUB(\varphi) \\
                      1,              & \text{otherwise, if } a \not \in D' \text{ or } b \in D'                                                                            \\
                      0,              & \text{otherwise, if } a \in D' \text{ and } b \not \in D'                                                                           \\
                  \end{cases}
              $
        \item $
                  a\tilde{\id}'b = TODO \begin{cases}
                      a\tilde{\id}b , & \text{if } a = v(\psi), b = v(\vartheta) \text{ and } \psi \id \vartheta \in \SUB(\varphi) \\
                      1,              & \text{otherwise, if } a = b,                                                               \\
                      0,              & \text{otherwise.}                                                                          \\
                  \end{cases}
              $
        \item $V'(\psi) = \begin{cases}
                      V(\psi) , & \text{if } \psi \in \SUB(\varphi)                                      \\
                      1,        & \text{if } \psi \not \in \SUB(\varphi) \text{ and } V(\psi) \in D      \\
                      0,        & \text{if } \psi \not \in \SUB(\varphi) \text{ and } V(\psi) \not \in D \\
                  \end{cases}$
    \end{itemize}

    We need to show that $\mathcal{M'}$ is an \SCI-model, $V'$ is a valuation and
    that $V'(\varphi) = V(\varphi)$:
    \begin{itemize}
        \item $\emptyset \not = U' \supsetneq D' \not = \emptyset$ -- trivially from $0, 1 \in U', 1 \in D'$.
        \item $\tilde{\lnot}'a \in D'$ if and only if $a \not \in D'$ -- there are cases:
              \begin{enumerate}
                  \item[1°] There exists $\psi$, such that $a = V(\psi)$ and $\lnot \psi \in \SUB(\varphi)$. Then from the definition of $\tilde{\lnot}'$ we have $\tilde{\lnot}'a = V'(\lnot \psi)$, from the definition of $V'$ we have that $V'(\lnot \psi) = V(\lnot \psi)$ and since $V$ is a valuation we have that $V(\lnot \psi) = \tilde{\lnot}V(\psi)$. But we assumed that $a = V(\psi)$, so we have that $\tilde{\lnot}'a = \tilde{\lnot}a$. Then since $\mathcal{M}$ is a model, we know that $\tilde{\lnot}a \in D$ if and only if $a \not \in D$, and since $\lnot\psi \in \SUB(\varphi)$, we have $\tilde{\lnot}'a \in D'$ if and only if $a \not \in D'$.
                  \item[2°] Otherwise, and $a \in D'$. Then $\tilde{\lnot}'a = 0 \not \in D'$.
                  \item[3°] Otherwise, and $a \not \in D'$. Then $\tilde{\lnot}'a = 1 \in D'$.
              \end{enumerate}
        \item For all $\psi \in \FOR$, $V'(\lnot \psi) = \tilde{\lnot}'V'(\psi)$ -- there are
              cases:
              \begin{enumerate}
                  \item[1°] $\lnot \psi \in \SUB(\varphi)$. Then from the definition of $\tilde{\lnot}'$ we have that $\tilde{\lnot}'V'(\psi) = V'(\lnot \psi)$.
                  \item[2°] $\lnot \psi \not \in \SUB(\varphi)$.

                          {\color{red} Kontrprzykład:
                              \begin{itemize}
                                  \item weźmy $\psi$: $\lnot \psi \not \in \SUB(\varphi)$
                                  \item Z definicji $V'$: $V'(\lnot \psi) = 1 $ lub $0$, w zeleżności czy $V(\psi) \in
                                            D$.
                                  \item Niech $V'(\psi) = b$
                                  \item Niech istnieje $\chi$, takie że: $\lnot \chi \in \SUB(\varphi)$, $V'(\chi) =
                                            b$, $V'(\lnot \chi) = c \not \in \{1, 0\}$
                                  \item wtedy, z definicji $\tilde{\lnot}'V'(\psi) = \tilde{\lnot}'b = c \not \in \{1,
                                            0\}.$ Sprzeczność.
                              \end{itemize}

                              Konkretny kontrprzykład:
                              \begin{itemize}
                                  \item $\varphi = \lnot(p \id \lnot p)$
                                  \item $U = \{a, b\}$
                                  \item $D = \{b\}$
                                  \item $V(p) = V(p \id \lnot p) = a$
                                  \item $V(\lnot p) = V(\lnot(p \id \lnot p)) = b$
                                  \item Konstruujemy $\M',V'$ jak wyżej
                                  \item $\psi := \lnot(p \id \lnot p)$
                                  \item $V'(\lnot \psi) = 0$, bo $\lnot \psi \not \in \SUB(\varphi)$
                                  \item $V'(\psi) = a$
                                  \item $\tilde{\lnot}'a = \tilde{\lnot}a = b$, bo istnieje $\chi = p$ t.że $a=V(\chi)$ i $\lnot \chi \in \SUB(\varphi)$
                                  \item więc $\tilde{\lnot}'V'(\psi) = b \not = 0$ -- sprzeczność.
                              \end{itemize}

                              Trzeba chyba skomplikować definicję $V'(\psi)$ rozbijając $\psi$. Np. jeśli
                              $\exists \chi: \lnot \chi = \psi, \chi \in \SUB(\varphi)$ to $V'(\psi) =
                                  \tilde{\lnot}'V'(\chi)$. Wtedy $V'$ przestaje mieć tą ładną właściwość, że
                              $\psi \not \in \SUB(\varphi) \Rightarrow V'(\psi) \in \{1, 0\}$, np. w
                              przykładzie u góry $V'(\lnot \lnot (p \id \lnot p))$ będzie równe $a$. Ale nie
                              ma to wpływu na liczność modelu i wydaje mi się że będzie się wszystko zgadzać.
                          }
              \end{enumerate}
        \item $a \tilde{\ra}'b \in D'$ if and only if $a \not \in D'$ or $b \in D'$ -- there are cases:
              \begin{enumerate}
                  \item[1°] TODO There exists $\psi$ and $\vartheta$ such that $a = v(\psi), b=v(\vartheta)$ and $\psi \ra \vartheta \in \SUB(\varphi)$. Then from the definition of $\tilde{\ra}'$ we have that $a \tilde{\ra}'b \in D' \iff a \tilde{\ra} b \in D$, from the definition of $\tilde{\ra}$ we have that $a \tilde{\ra} b \in D$ if and only if $a \not \in D$ or $b \in D$, which from the definition of $D'$ is if and only if $a \not \in D'$ or $b \in D'$.
                  \item[2°] TODO There are no such $\psi$ and $\vartheta$ and $a \not \in D$. Then, from the definition of $\tilde{\ra}'$ we have that $a \tilde{\ra}' b = 1 \in D'$ and from the definition of $D'$ we have that $a \not \in D'$.
                  \item[3°] TODO There are no such $\psi$ and $\vartheta$ and $b \in D$. Then again, from the definition of $\tilde{\ra}'$ we have that $a \tilde{\ra}' b = 1 \in D'$ and from the definition of $D'$ we have that $b \in D'$.
              \end{enumerate}
    \end{itemize}
\end{proof}

\begin{theorem}[Decidability of \SCI]
    \label{decidability}
    \SCI is decidable.
\end{theorem}
\begin{proof}
    Let us take any \SCI-formula $\varphi$, we want to know if $\varphi$ is a \SCI-theorem. From fullness and completeness of \SCI we know that $\varphi$ is a theorem if and only if $\lnot \varphi$ has no \SCI-model and valuation $\M, V$, such that $\M, V \models \lnot \varphi$. But from \cref{finite_model} we know that if there is $\M, V$, such that $\M, V \models \lnot \varphi$, then there is a finite model $\M'$ and $V'$ such that $\M', V' \models \lnot \varphi$. Furthermore, from the proof of \cref{finite_model} we can see that $|U'| \leq |\SUB(\varphi)|$ ($|X|$ here meaning the size of a set $X$).

    For any finite model $\M'$ we can construct an isomorphic model $\M''$ with
    $U'' = \{1, 2, ..., |U'|\}$ by setting any bijection $f : U' \rightarrow U''$
    and then setting the rest of the model and $V''$ accordingly. So, we know that
    if there is $\M, V$ such that $\M, V \models \lnot \varphi$, then and only then
    there is a $\M'', V''$, such that $\M'', V'' \models \lnot \varphi$ and $U'' =
        \{1, 2, ..., |\SUB(\lnot \varphi)|\}$, and there is a finite amount of such
    finite models.

    We can have a procedure to set $D''$ to be any of $2^{|\SUB(\lnot \varphi)|}$
    subsets of $U''$, then for any $a \in U''$ to set $\tilde{\lnot}a$ to be any
    element of $U''$, then for any $a, b \in U''$ to set $a \tilde{\ra} b$ to be
    any element of $U''$, then for any $a, b \in U''$ to set $a \tilde{\id} b$ to
    be any element of $U''$ and then to set every variable of $\lnot \varphi$ to be
    any element of $U''$ and finally to check, whether such constructed $\M'', V''$
    is a \SCI-model and a correct valuation. If it happens to be the case, then we
    have found a \SCI-model invalidating $\varphi$. If after checking every such
    combination we have not found any \SCI-model invalidating $\varphi$, we know
    that $\varphi$ is valid, and so that it is a theorem.

    If we set $n = s(\varphi)$, then this procedure will take $O({n^2} ^ {n^2})$
    operations, so it is very far from practical, but it is nevetheless finite, so
    we have a finite procedure to decide whether $\varphi$ is a \SCI-theorem.

\end{proof}

\section{Deduction in \SCI}

\subsection{Deduction systems}

The axiomatization of \SCI we presented in \cref{axiomatization} is an example
of a Hilbert-style deductive system. Such systems consist of a set of axioms
and inference rules (often \emph{modus ponens} is the only inference rule) that
can be used to deduce theses, as explained in \cref{thesis}.

Hilbert-style systems are historically first and most fundamental deductive
systems. For every thesis of such systems, there is a sequence of formulas all
the way back to the axioms, from which a resulting thesis is inferred. But
trying to prove some specific formula can be very challenging as it requires
creative usages of the axioms so that the given formula is inferred at the end.
Those proof systems give no clue how to do that.

The same is true for the deduction system presented in
\cref{axiomatization,thesis}. What is more, from \cref{decidability} we know
that \SCI is decidable, but so far we have no practical decision procedure to
determine whether a given formula is a tautology.

Let us show some examples to illustrate these points:

\begin{example}
    Let us show a full proof of a formula: $(p \id q) \ra (q \id p)$

    \begin{enumerate}
        \item $p \id p$ \hfill $(Ax1)$
        \item $(p \id q) \ra ((p \id p) \ra ((p \id p) \id (q \id p)))$ \hfill $(Ax5)$
        \item $((p \id p) \id (q \id p)) \ra ((p \id p) \ra (q \id p))$ \hfill $(Ax2)$
        \item $(p \id p) \ra (((p \id q) \ra ((p \id p) \ra ((p \id p) \id (q \id p)))) \ra ((((p \id p) \id (q \id p)) \ra ((p \id p) \ra (q \id p))) \ra ((p \id q) \ra (q \id p))))$
              \begin{flushright}
                  \PC axiom $\varphi \ra ((\psi \ra (\varphi \ra \vartheta)) \ra ((\vartheta \ra (\varphi \ra \chi)) \ra (\psi \ra \chi)))$
                  \\
                  with $\varphi = p \id p$\\$\psi = p \id q$\\$\vartheta = q \id p$\\$\chi = (p \id p) \id (q \id p)$.
              \end{flushright}
        \item $((p \id q) \ra ((p \id p) \ra ((p \id p) \id (q \id p)))) \ra ((((p \id p) \id (q \id p)) \ra ((p \id p) \ra (q \id p))) \ra ((p \id q) \ra (q \id p)))$ \hfill MP(1, 4)
        \item $(((p \id p) \id (q \id p)) \ra ((p \id p) \ra (q \id p))) \ra ((p \id q) \ra (q \id p))$ \hfill MP(2, 5)
        \item $(p \id q) \ra (q \id p)$ \hfill MP(3, 6)
    \end{enumerate}

\end{example}
As can be seen, althouth the finished proof isn't very long, it would be very
difficult to come up with the precise axioms to use.

\begin{example}
    Let us take a formula $p \id q$. We could be deducing many theses from our axioms, but at no point will we deduce it, nor $\lnot(p \id q)$, because those formulas aren't tautologies. But we cannot arrive at this conclusion using just our Hilbert-style system.
\end{example}

\begin{example}
    Let us show a full proof of a formula $(q \id r) \ra ((p \id q) \ra (p \id r))$.

    \begin{enumerate}
        \item $(((p \id q) \id (p \id r)) \ra ((p \id q) \ra (p \id r))) \ra ((q \id r) \ra (((p \id q) \id (p \id r)) \ra ((p \id q) \ra (p \id r)))) $
              \begin{flushright} \PC axiom $\varphi \ra (\psi \ra \varphi)$ \\
                  with $\varphi = ((p \id q) \id (p \id r)) \ra ((p \id q) \ra (p \id r))$\\
                  $\psi = q \id r$
              \end{flushright}
        \item $((p \id q) \id (p \id r)) \ra ((p \id q) \ra (p \id r))  $ \hfill $(Ax2)$
        \item $(q \id r) \ra (((p \id q) \id (p \id r)) \ra ((p \id q) \ra (p \id r)))$ \hfill MP(2, 1)
        \item $((q \id r) \ra (((p \id q) \id (p \id r)) \ra ((p \id q) \ra (p \id r)))) \ra (((q \id r) \ra ((p \id q) \id (p \id r))) \ra ((q \id r) \ra ((p \id q) \ra (p \id r)))) $ \begin{flushright}
                  \PC axiom $(\varphi \ra (\psi \ra \vartheta)) \ra ((\varphi \ra \psi) \ra (\varphi \ra \vartheta))$ \\
                  with $\varphi = q \id r$ \\
                  $\psi = (p \id q) \id (p \id r)$ \\
                  $\vartheta = (p \id q) \ra (p \id r)$
              \end{flushright}
        \item $((q \id r) \ra ((p \id q) \id (p \id r))) \ra ((q \id r) \ra ((p \id q) \ra (p \id r))) $ \hfill MP(3, 4)
        \item $p \id p $ \hfill $(Ax1)$
        \item $(p \id p) \ra ((q \id r) \ra ((p \id q) \id (p \id r)))$ \hfill $(Ax5)$
        \item $(q \id r) \ra ((p \id q) \id (p \id r))$ \hfill MP(6, 7)
        \item $(q \id r) \ra ((p \id q) \ra (p \id r))$ \hfill MP(8, 5)
    \end{enumerate}
\end{example}
% ------------------------------------------------------------------

% \begin{definition}[Complexity of a set of formulas]
%     Given a set of formulas $\Phi$, let us define:
%     $$
%         c(\Phi) = \sum_{\phi \in \Phi} 3^{s(\phi)}
%     $$
% \end{definition}

% Let us assume that we are given a full ordering of formulas $\prec$, such that:
% \begin{itemize}
%     \item if $\varphi \not = \psi$, then $\varphi \prec \psi$ or $\psi \prec \varphi$,
%     \item if $s(\varphi) < s(\psi)$, then $\varphi \prec \psi$,
%     \item if $\varphi \prec \psi$, then $\varphi \id \psi \prec \psi \id \varphi$.
% \end{itemize}

% \section{Satisfiability in \SCI}
% \subsection{\SCI is decidable}
% \subsection{Decision procedures}
% \subsubsection{$T$ decision procedure}
% \subsubsection{$\Ts$ decision procedure}

% \leqnomode
% \begin{figure}
%     \begin{subfigure}{\textwidth}
%         \centering
%         \begin{subfigure}{0.3\textwidth}
%             \begin{equation}
%                 \tag{$\lnot$}
%                 \frac{
%                     \Phi \cup \{ \lnot \lnot \varphi \}}%
%                 { \Phi \cup \{\varphi\} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.3\textwidth}
%             \begin{equation}
%                 \tag{$\rightarrow$}
%                 \frac{
%                     \Phi \cup \{ \varphi \rightarrow \psi \}}%
%                 { \Phi \cup \{ \lnot \varphi \} | \Phi \cup \{ \psi \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.3\textwidth}
%             \begin{equation}
%                 \tag{$\lnot \rightarrow$}
%                 \frac{
%                     \Phi \cup \{\lnot ( \varphi \rightarrow \psi ) \}}%
%                 { \Phi \cup \{ \varphi, \lnot \psi \} }
%             \end{equation}
%         \end{subfigure}
%         \caption{Decomposition rules}
%     \end{subfigure}
%     \begin{subfigure}{\textwidth}
%         \centering
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$\texttt{sym}_1^1$}
%                 \frac{
%                     \Phi \cup \{ \varphi \equiv \psi \} }%
%                 { \Phi \cup \{\psi \equiv \varphi\} }
%                 \text{ if $\varphi$ > $\psi$}
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$\texttt{sym}_2^1$}
%                 \frac{
%                     \Phi \cup \{ \varphi \not \equiv \psi \}}%
%                 { \Phi \cup \{\psi \not \equiv \varphi\} }
%                 \text{ if $\varphi$ > $\psi$}
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.5\textwidth}
%             \begin{equation}
%                 \tag{$\texttt{fun}^1$}
%                 \frac{
%                     \Phi \cup \Phi(\varphi) \cup \{ p \equiv \varphi \} }%
%                 { \Phi \cup \Phi(\varphi / p) \cup \{p \equiv \varphi\} }
%                 \text{ if $p$ < $\varphi$}
%             \end{equation}
%         \end{subfigure}
%         \caption{Reduction rules}
%     \end{subfigure}
%     \begin{subfigure}{\textwidth}
%         \centering
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$\not \equiv_1$}
%                 \frac{
%                     \Phi \cup \{ p \not \equiv \psi \} }%
%                 { \Phi \cup \{ p \not \equiv \alpha, \alpha \equiv \psi \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$\not \eq_2$}
%                 \frac{
%                     \Phi \cup \{ \varphi \not \id \psi \} }%
%                 { \Phi \cup \{ \alpha \id \varphi, \beta \id \psi, \alpha \not \id \beta \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$\eq_\lnot$}
%                 \frac{
%                     \Phi \cup \{ p \id \lnot \psi  \} }%
%                 { \Phi \cup \{ p \id \lnot \alpha, \alpha \id \psi \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.5\textwidth}
%             \begin{equation}
%                 \tag{$ \eq_\ra $}
%                 \frac{
%                     \Phi \cup \{ p \id (\varphi \ra \psi) \} }%
%                 { \Phi \cup \{ p \id (\alpha \ra \beta), \alpha \id \varphi, \beta \id \psi \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$ \eq_\ra^l $}
%                 \frac{
%                     \Phi \cup \{ p \id (q \ra \psi) \} }%
%                 { \Phi \cup \{ p \id (q \ra \alpha), \alpha \id \psi  \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$ \eq_\ra^r $}
%                 \frac{
%                     \Phi \cup \{ p \id (\varphi \ra q) \} }%
%                 { \Phi \cup \{ p \id (\alpha \ra q), \alpha \id \varphi  \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$ \eq_\id^l $}
%                 \frac{
%                     \Phi \cup \{ p \id (q \id \psi) \} }%
%                 { \Phi \cup \{ p \id (q \id \alpha), \alpha \id \psi  \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$ \eq_\id^r$}
%                 \frac{
%                     \Phi \cup \{ p \id (\varphi \id q) \} }%
%                 { \Phi \cup \{ p \id (\alpha \id q), \alpha \id \varphi \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.45\textwidth}
%             \begin{equation}
%                 \tag{$ \eq_\id $}
%                 \frac{
%                     \Phi \cup \{ p \id (\varphi \id \psi) \} }%
%                 { \Phi \cup \{ p \id (\alpha \id \beta), \alpha \id \varphi, \beta \id \psi \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$ \id$}
%                 \frac{
%                     \Phi \cup \{ \varphi \id \psi \} }%
%                 { \Phi \cup \{ \alpha \id \beta, \alpha \id \varphi, \beta \id \psi \} }
%             \end{equation}
%         \end{subfigure}
%         \caption{Identity rules. Note: in all identity rules the highlighted formula is not simple (i.e. $\varphi$ and $\psi$ are not variables)}
%     \end{subfigure}
%     \caption{Deduction rules of the first phase of $\Ts$.}
%     \label{fig:rules1}
% \end{figure}

% \begin{figure}
%     \begin{subfigure}{\textwidth}
%         \centering
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$\texttt{sym}_1^2$}
%                 \frac{
%                     \Phi \cup \{ \varphi \equiv \psi \} }%
%                 { \Phi \cup \{\psi \equiv \varphi, \varphi \id \psi \} }
%                 \text{ if $\varphi$ > $\psi$}
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.4\textwidth}
%             \begin{equation}
%                 \tag{$\texttt{sym}_2^2$}
%                 \frac{
%                     \Phi \cup \{ \varphi \not \equiv \psi \}}%
%                 { \Phi \cup \{\psi \not \equiv \varphi, \varphi \not \id \psi\} }
%                 \text{ if $\varphi$ > $\psi$}
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.55\textwidth}
%             \begin{equation}
%                 \tag{$\texttt{fun}^2$}
%                 \frac{
%                     \Phi \cup \Phi(\varphi) \cup \{ p \equiv \varphi \} }%
%                 { \Phi \cup \Phi(\varphi / p) \cup \Phi(\varphi) \cup \{p \equiv \varphi\} }
%                 \text{ if $p$ < $\varphi$}
%             \end{equation}
%         \end{subfigure}
%     \end{subfigure}
%     \begin{subfigure}{\textwidth}
%         \begin{subfigure}{0.48\textwidth}
%             \begin{equation}
%                 \tag{$ \eq_\top $}
%                 \frac{
%                     \Phi \cup \{ p \id q \} }%
%                 { \Phi \cup \{ p, q, p\id q \} | \Phi \cup \{ \lnot p, \lnot q, p\id q \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{subfigure}{0.48\textwidth}
%             \begin{equation}
%                 \tag{$ \eq_\top^\lnot $}
%                 \frac{
%                     \Phi \cup \{ p \id \lnot q \} }%
%                 { \Phi \cup \{ p, \lnot q, p \id \lnot q \} | \Phi \cup \{ \lnot p, q, p\id \lnot q \} }
%             \end{equation}
%         \end{subfigure}
%         \begin{equation}
%             \tag{$ \eq_\top^\ra $}
%             \frac{
%                 \Phi \cup \{ p \id (q \ra r) \} }%
%             { \Phi \cup \{ p, \lnot q, \theta \} | \Phi \cup \{ p, r, \theta \} | \Phi \cup \{ \lnot p, q, \lnot r, \theta \}}
%             \text{, where } \theta = (p \id (q \ra r))
%         \end{equation}
%         \begin{equation}
%             \tag{$ \eq_\top^\id $}
%             \frac{
%                 \Phi \cup \{ p \id (q \id r) \} }%
%             { \Phi \cup \Phi_1 |
%                 \Phi \cup \Phi_2 |
%                 \Phi \cup \Phi_3 |
%                 \Phi \cup \Phi_4 |
%                 \Phi \cup \Phi_5 |
%                 \Phi \cup \Phi_6
%             }
%             \text{, where }
%         \end{equation}
%         \[
%             \begin{aligned}
%                 \Phi_1 & = \{p, q, r, q \id r, p \id (q \id r) \}                       \\
%                 \Phi_2 & = \{p, \lnot q, \lnot r, q \id r, p \id (q \id r) \}           \\
%                 \Phi_3 & = \{\lnot p, q, r, q \not \id r, p \id (q \id r) \}            \\
%                 \Phi_4 & = \{\lnot p, q, \lnot r, q \not \id r, p \id (q \id r) \}      \\
%                 \Phi_5 & = \{\lnot p,\lnot q, r, q \not \id r, p \id (q \id r) \}       \\
%                 \Phi_6 & = \{\lnot p, \lnot q,\lnot r, q \not \id r, p \id (q \id r) \} \\
%             \end{aligned}
%         \]
%         \begin{equation}
%             \tag{$\eq_\bot$}
%             \frac{
%                 \Phi \cup \{ p \not \id q \} }%
%             { \Phi \cup \{p,q,\theta\} |
%                 \Phi \cup \{p,\lnot q,\theta\} |
%                 \Phi \cup \{\lnot p,q,\theta\} |
%                 \Phi \cup \{\lnot p,\lnot q,\theta\}
%             }
%             \text{, where } \theta = (p \not \id q)
%         \end{equation}
%     \end{subfigure}
%     \caption{Deduction rules of the second phase of $\Ts$.}
%     \label{fig:rules2}
% \end{figure}

% The $\Ts$ decision procedure has two main groups of deduction rules, see figures \ref{fig:rules1} and \ref{fig:rules2}.

% The rules are

% TODO: differences compared to \enquote{Dedukcyjne Dylematy} (DD):
% \begin{itemize}
%     \item Phase 1 rules replace formulas, phase 2 rules only add formulas.
%     \item No need to say: \enquote{w wyniku zastosowania reguły co najmniej jedna z konkluzji zawiera formułę, która nie występuje w $X \cup \{\varphi\}$}. In the first phase simply apply all the rules as long as you can. In second phase apply closure rules only once to any given formula.
%     \item There are TWO $fun$ rules: one for each phase. Phase 1 fun rule replaces formulas, phase 2 fun rule only adds formulas (leaves the original unchaged as well). If $fun$ is replacing in phase 2 then e.g. after rule $\equiv_T^\lnot$ the rule $fun$ is applied removing one of the conclusion formulas. Besides, the examples in DD seem to assume the $fun$ works this way anyways.
% \end{itemize}

% \subsubsection{Correctness and completeness of $\Ts$}
% \begin{theorem}
%     $\Ts$ is correct.
% \end{theorem}

% \begin{proof}
%     TODO: this is just a sketch, rewrite. We want to show that if $\Ts$ says the formula is closed, it is indeed closed.
%     \begin{enumerate}
%         \item All the rules of $\Ts$ are strongly correct. TODO at least the non-obvious ones.
%         \item We're given a closed tree.
%         \item Going up the tree, on every step we have a closed set of formulas, because every rule is strongly correct.
%         \item The root is closed.
%     \end{enumerate}
% \end{proof}

% \begin{theorem}
%     $\Ts$ is complete.
% \end{theorem}

% We want to show, that if the formula is closed, the tree produced by $\Ts$ is closed, or by contraposition, that if the tree produced by $\Ts$ is open, the given formula is open.

% First, let us show that the first phase finishes.

% \begin{theorem}
%     The first phase of $\Ts$ finishes and produces finite trees.
% \end{theorem}
% \begin{proof}
%     Let us first notice that all decomposition and identity rules of the first phase reduce the complexity of the formulas.

%     \begin{lemma}
%         \label{l:c}
%         If rule $R$ is a decomposition rule or an identity rule, $\Phi$ is its premise and $\Phi_1, ... \Phi_n$ are its conclusions, then $c(\Phi) > c(\Phi_i)$, for $1 \leq i \leq n$.
%     \end{lemma}
%     \begin{proof}
%         TODO: Przerachować. Chyba zdaje się działać, ewentualnie trochę podkręcić $s$ lub $c$.
%     \end{proof}
%     \begin{lemma}
%         \label{l:c2}
%         If rule $R$ is a reduction rule $\Phi$ is its premise and $\Phi_1$ is its conclusion, then $c(\Phi) \geq c(\Phi_1)$.
%     \end{lemma}
%     \begin{proof}
%         Trivial. All the reduction rule replace a formula with another formula that is smaller in our given order $\prec$, so they will never increase the complexity function.
%     \end{proof}
%     \begin{lemma}
%         \label{l:c3}
%         The reduction rules cannot be applied indefinitely without lowering the complexity function.
%     \end{lemma}
%     \begin{proof}
%         TODO: przepisać. To jest trywialne.

%         Based on lemma \cref{l:c2}, applying the reduction rule will never increase the complexity function. If applying the rule will give conclusions with the same complexity, the conclusion will contain some formulas that are lower in $\prec$ than the appropriate formulas in the premise, while all the other formulas are equal.
%     \end{proof}

%     As a consequence of lemmas \ref{l:c}, \ref{l:c2} and \ref{l:c3} it is clear that, for each node of the produced tree, after a finite number of rules applied, either a closed node or a node with lower complexity will be reached. Since the complexity of the node is always positive and finite it is impossible that the complexity will be lowered indefinitely, so the first phase of the $\Ts$ procedure will finish.

% \end{proof}

% \begin{theorem}
%     All leaves produced by the first phase are either closed or contain only simple formulas.
% \end{theorem}
% \begin{proof}
%     Simple. Assume there is a non-simple formula in an open leaf. One of decomposition, reduction or equality rules would apply to it, which is a contradiction with the fact that the first phase finished.
% \end{proof}

% \begin{theorem}
%     Given a set of simple formulas, phase 2 finishes.
% \end{theorem}
% \begin{proof}
%     TODO: trywialne. Wszystkie formuły z $\id$ w nazwie wykonają się co najwyżej raz, po każdej z nich $\texttt{sym}*$ i $\texttt{fun}$ też co najwyżej raz.
% \end{proof}

% Let us call a tree that is produced by the $\Ts$ procedure a $\Ts$ tree.

% \begin{theorem}
%     Open leaf $X$ of a $\Ts$ tree satisfies the following conditions:
%     \begin{itemize}
%         \item [(X1)] $X$ is a set of simple formulas.
%         \item [(X2)] If $\varphi \in \FOR^X$ then $\varphi$ is either a simple formula or is equal to $p \ra q$, for some variables $p, q$.
%         \item [(X3)] If $\varphi \id \psi \in \FOR^X$, then $\varphi \leq \psi$ or $\psi \id \varphi \in \FOR^X$.
%         \item [(X4)] If $\varphi \not \id \psi \in \FOR^X$, then $\varphi < \psi$ or $\psi \not \id \varphi \in \FOR^X$.
%         \item [(X5)] If $p \id q \in X$, then $p, q \in X$, or $\lnot p, \lnot q \in X$.
%         \item [(X6)] If $p \id \lnot q \in X$, then $p, \lnot q \in X$, or $\lnot p, q \in X$.
%         \item [(X7)] If $p \id (q \ra r) \in X$, then $p, \lnot q \in X$, or $p, r \in X$, or $\lnot p, q, \lnot r \in X$.
%         \item [(X8)] If $p \id (q \id r) \in X$, then $\Phi_i \subseteq X$, for $i \in \{1, ..., 6\}$, where $\Phi_i$ are defined in figure \ref{fig:rules2}.
%     \end{itemize}
% \end{theorem}
% \begin{proof}
%     (X1) Phase 1 ends with nodes containing only simple formulas. Phase 2 doesn't add any non-simple formulas.

%     (X2), ... (X8) Trivial, directly following phase 2 rules.
% \end{proof}

% \begin{lemma}[Lemat 5 z DD]
%     Let $X$ be an open leaf of a $\Ts$ tree and let $\lnot \varphi \in \FOR^X$. Then:
%     $$
%         \varphi \in X \cup \IDx \text{ iff } \lnot \varphi \not \in X \cup \IDx.
%     $$
% \end{lemma}
% \begin{proof}
%     Because of $X2$, if $\lnot \varphi \in \FORx$, then $\varphi = p$ or $\varphi = (p \id q)$ for some variables $p, q$.

%     \begin{itemize}
%         \item [1°] $\varphi = p$. ... (Jak w DD.)
%         \item [2°] $\varphi = (p \id q)$. \\
%               Since $\lnot \varphi = p \not \id q \in \FORx$, we have that $p \not \id q \in X$, because $p \not \id q$ cannot be a subformula of a simple formula. From $p \not \id q \in X$ it follows that $p \not = q$ and $p \id q \not \in X$, otherwise $X$ would be an axiom set and therefore closed.

%               So, we have: $p \not \id q \in X$ (so, $\lnot \varphi \in X \subseteq X \cup \IDx$). We need to show that $\varphi \not \in X \cup \IDx$, i.e. $p \id q \not \in X \cup \IDx$. We have that $p \id q \not \in X$, so all we need to show is that $p \id q \not \in \IDx$.

%               Let as assume otherwise, that $p \id q \in \IDx$.

%               Since $p \id q \not \in X$, we have that $p \id q \not \in \IDx_0$. $p \id q \not \in \IDx_n$ for $n > 0$ based on the definition of $\IDx_n$. So, for the $p \id q$ to be in $\IDx$ we need to have formulas $\theta_1, ..., \theta_k$, for some $k \geq 1$, such that: $p \id \theta_1 \in X$, or $\theta_1 \id p \in X$, ..., $\theta_k \id q$, or $q \id \theta_k$.
%               \begin{itemize}
%                   \item [2.1°] $k = 1$.\\
%                         TODO: przepisać żeby było jaśniej.\\
%                         We have that $p \id \theta_1 \in X$, or $\theta_1 \id p \in X$ and $\theta_1 \id q \in X$, or $q \id \theta_1 \in X$. One of the formulas in $\{p, q, \theta_1\}$ will be smallest in the ordering $\prec$. If the formula $p$ is the smallest, the $\texttt{fun}$ rule would produce $q \id p \in X$ (or $p \id q \in X$) from $q \id \theta_1 \in X$ (or $\theta_1 \id q \in X$) by replacing $\theta_1$ by $p$. In any case, $p \id q$ would be in $X$ (possibly after $\texttt{sym}^2_1$). Thus $X$ would be an axiom set ($p\not \id q, p \id q \in X$), a contradiction. A contradiction happens in a similar way if the formula $q$ is the smallest of the three. If the formula $\theta_1$ is the smallest one, from $p \not \id q \in X$ will be produced either $\theta_1 \not \id q \in X$ or $p \not \id \theta_1 \in X$ (by $\texttt{fun}$ similarly to above), giving another contradiction.
%                   \item [2.2°] $k > 1$.\\
%                         ???
%               \end{itemize}
%     \end{itemize}
% \end{proof}

% \section{Implementation}
% \subsection{Data structures}
% \subsection{Optimizations}
% % \section{Comparison of $T$ and $T*$ systems}
% \subsection{Size of produced proof trees}
% \subsection{Time complexity}
% \subsection{Real time}

\begin{thebibliography}{9}
    \bibitem{suszko_bloom}
    Stephen L. Bloom, Roman Suszko (1972) \emph{Investigations into the Sentential Calculus with Identity}, Notre Dame Journal of Formal Logic, vol. XIII, no 3.

    % \bibitem{niebieska}
    % Zdzisław Augustynek i Jacek Juliusz Jadacki (1993) \emph{Possible Ontologies}, \emph{Poznań Studies in the Philosophy of the Sciences and the Humanities}, tom 29, wyd. Rodopi.

    % \bibitem{wazniak}
    % \emph{Logika i teoria mnogości}, \url{https://wazniak.mimuw.edu.pl/index.php?title=Logika_i_teoria_mnogo%C5%9Bci}.

    % \bibitem{gowers}
    % William Timothy Gowers  \emph{How to use Zorn’s lemma}, \url{https://gowers.wordpress.com/2008/08/12/how-to-use-zorns-lemma/}
\end{thebibliography}

\end{document}